Neutrino Annihilation Around Massive Nuclear Accretion Disks

### 2015-11-10 paper.tex
Add paper boilerplate

\documentclass[aps,floatfix,prd,superscriptaddress,twocolumn]{revtex4-1}
% Intended to be built with pdflatex, using the REVTEX 4.1 package.
% Template taken from FermiLab APS Journal template.

% packages
\usepackage{amssymb}              % for \gtrsim ...
\usepackage{amsmath}              % for \text, align, split ...
\usepackage{color}                % for colorful margin notes
\usepackage[mathscr]{euscript}    % for \mathscr
\usepackage{float}                % for better control of figure placement
\usepackage{graphicx}             % for \includegraphics
\usepackage{hyperref}             % for \url and \href
\usepackage{listings}             % for typesetting code with \lstinline
\usepackage{tabularx}             % for tabularx

% commands/initialization
\hyphenation{ALPGEN} % avoid incorrect hyphenation
\hyphenation{EVTGEN} % ''
\hyphenation{PYTHIA} % ''
\lstset{basicstyle=\ttfamily} % define \lstinline for inline code snippet

\newcommand{\todo}[1]{\marginpar{\tiny{\textcolor{red}{#1}}}}
%\renewcommand\todo[1]{} % if you wish to hide all todo notes

% force all figures to be placed right where they are in the text
\floatplacement{figure}{H}

@@

\widetext
\leftline{Compiled \today}
\leftline{To be submitted to PRD}

\title{Elastic Scattering in General Relativistic Ray Tracing for Neutrinos}

\author{M.\ Brett Deaton}
\email{mbdeaton@ncsu.edu}
\affiliation{Joint Institute for Nuclear Astrophysics,
  Michigan State University, East Lansing, MI 48824, USA}
\affiliation{Department of Physics,
  North Carolina State University, Raleigh, NC 27695, USA}

@@

\author{Jerred Jesse}
\affiliation{Department of Physics \& Astronomy,
  Washington State University, Pullman, WA 99164, USA}

\author{G.\ C.\ McLaughlin}
\affiliation{Department of Physics,
  North Carolina State University, Raleigh, NC 27695, USA}

% *******************************************************************************

\begin{abstract}
  We present a covariant ray tracing algorithm for computing high-resolution
  neutrino distributions in general numerical spacetimes with hydrodynamical
  sources.
  Our formulation treats the very important effect of
  elastic scattering of neutrinos off of nuclei and nucleons
  (changing the neutrino's direction but not energy)
  by incorporating estimates of the background neutrino fields.
  These background fields may be taken from a low-order moment simulation
  or be ignored entirely, in which case the method
  reduces to a standard state-of-the-art ray tracing formulation.
  It handles radiation in regimes spanning optically thick to optically thin.
  We test the new code, highlight its strengths and weaknesses, and
  apply it to simulations of neutron star mergers to
  compute neutrino fluxes as a function of energy and emission angle.
\end{abstract}

@@

\section{Introduction}
Neutrinos are one of the dominant energy transport phenomena at play in
neutron star mergers, heating, cooling, and pushing the disrupted nuclear
matter.
In addition, they change the composition of the matter via charged current
interactions.
Because neutrinos scatter over length scales both large and small with
respect to fluid scales, accurate models require a neutrino treatment that
respects the freedom of neutrino distribution functions to vary
drastically from simple geometric distributions in thermodynamic equilibrium.

This is a challenging task in the environment of a merger,
which generally lacks any spatial symmetries so that fully general
seven-dimensional solutions to the Boltzmann Equation are not feasible.
Leakage approximations
\citep{deat2013-leakage, pere2016-asl, radi2016-dynamical}
\todo{cite others}
capture the qualitative effects of neutrinos on the matter, but
provide extremely limited information about the neutrino field itself.
\todo{discuss mc more clearly}
Monte Carlo methods \citep{abdi2012-monte_carlo, rich2015-monte_carlo}
are an excellent tool require large computational resources because of
their global treatment of the problem.

The state of the art today couples radiation and matter with a truncated
moment formalism
\citep{shib2011-truncated_moment, fouc2015-m1_nsbh, fouc2016-m1_nsns,
  just2015-m1_code},
\todo{cite others}
evolving the zeroth- and first-angular moments of the distribution function
(the energy density and flux respectively) representing the zeroth-energy moment
of the radiation field (the total energy).
This method, commonly called M1 transport, was formulated by
\cite{thor1981-truncated_moment, shib2011-truncated_moment}
and has been applied for example in \cite{ramp2002-truncated_moment,
  card2013-truncated_moment, fouc2015-m1_nsbh, ocon2015-gr1d_with_nu}.
Recently, \cite{fouc2016-m1_evolve_n} have expanded their M1 transport code
to also evolve the number densities of neutrinos,
providing total and average energies throughout the simulated volume.
Even so, with M1 transport codes only evolving two angular moments
and one or two energy moments,
we can extract only limited angular and spectral information
from the neutrino fields.

%\subsubsection*{What physics problems do we want to address?}
But many interesting unsolved problems require an accurate model of the neutrino
spectra and angular distributions.
With a model of the neutrino emission from a merger we can
1) examine neutrino effects on the nucleosynthesis of the ejected material
\citep{surm2011-nickel_56, robe2016-sph_nu_nucleo},
2) explore the rich flavor oscillation physics likely to occur
\citep{malk2012-mnr_1, malk2015-mnr_2, malk2016-mnr_3, zhu2016-mnr_nsns_remnant,
  vaan2016-uncovering_mnr},
3) improve closure relations used in M1 transport schemes
\citep{ramp2002-truncated_moment, shib2011-truncated_moment,
  card2013-truncated_moment, fouc2015-m1_nsbh, ocon2015-gr1d_with_nu}, and
4) study possible jet formation due to neutrino annihilation
\citep{ruff1999-nunubar_nsns, asan2000-nunubar, birk2007-nunubar,
  hari2010-gr_nunubar_collapsar, zala2011-nunubar, leng2014-nunubar}.

%\subsubsection*{Why ray tracing?}
In this work we present a ray tracing method to compute neutrino distribution
functions from state-of-the-art general relativistic M1 transport
hydrodynamics simulations.
We choose ray tracing because it is conceptually simple, numerically cheap,
and easily extends to high resolution in energy and angle.

With a ray tracing method we approach radiation transport from the perspective
of a single observer at a spacetime event $x_o^\alpha$.
Our goal is to compute the distribution function
$f^{\nu_\sigma}(x_o^\alpha;p_\beta)$, or the amount of neutrino radiation
of species $\nu_\sigma\in\{\nu_{\{e,\mu,\tau\}},\bar{\nu}_{\{e,\mu,\tau\}}\}$
with momentum $p_\beta$ impinging on $x_o^\alpha$.
To do so we trace a geodesic trajectory from $x_o^\alpha$ in the backwards
direction $-p_\beta$ to sample the incoming radiation along that line of sight.
By tracing a family of rays intersecting $x_o^\alpha$ we build up a
picture of the distribution function there.
And by sampling many observation points we construct a global picture
of $f^{\nu_\sigma}(x^\alpha;p_\beta)$.

The ray tracing framework
is conceptually simple because it solves the equation of radiation
transport (Eqn.~\ref{eqn:boltzmann}) along characteristics,
reducing it to a one-dimensional ordinary differential equation.
It is numerically cheap because it confines computations of
$f^{\nu_\sigma}(x_o^\alpha;p_\beta)$ to the past light-cone of $x_o^\alpha$,
with the history of that light-cone truncated at large optical depth.
\todo{state scaling $\mathscr{O}(n)$?}
And it easily extends to high resolution in energy and angle by simply
increasing the number of rays sampled.

%\subsubsection*{What is new and better about this ray tracing formulation?}
Several ray tracing formulations for radiation transport already exist.
Most formulations assume an analytic spacetime metric
\citep{birk2007-nunubar, hari2010-gr_nunubar_collapsar, kova2011-gr_ray_tracing}.
And many make the simplifying assumption of blackbody emission from a
neutrinosurface \citep{birk2007-nunubar, kova2011-gr_ray_tracing},
limiting them to equilibrium, optically thick configurations.
Current state-of-the-art ray tracing formulations avoid the assumption of
blackbody spectra by integrating a local emissivity along each geodesic
(e.g. \cite{hari2010-gr_nunubar_collapsar} for neutrinos and
\cite{youn2012-gr_radiative_transfer} for photons).
But no formulations to date account for the important scattering and pair
processes outlined in Tab.~\ref{tab:neutrino_processes}.

We build upon existing ray tracing formulations
by eschewing any assumptions about the spacetime geometry,
integrating local emissivities,
and including elastic scattering in the integration along each geodesic.
To incorporate scattering processes in our method, we employ estimates of the
lowest-order moment of the neutrino distribution function
computed in an M1 transport simulation.
If moments are not available either from
an M1 transport evolution or a trustworthy analytical estimate,
our method reduces to the current state-of-the-art ray tracing method,
neglecting scattering.

%\subsubsection*{Why does a covariant formulation matter?}
We formulate our ray tracing equations covariantly---free from
assumptions about the spacetime geometry or coordinates. This is essential
because we want to apply the method as a postprocessing step using
time snapshots of data computed from general relativistic evolutions.
The spacetime represented in these snapshots is not analytic (i.e. Kerr).
And even in configurations that are described well by the Kerr metric
(e.g. a low-mass disk around a massive black hole),
the evolution coordinates are unlikely to present the metric in
a familiar analytic form.
This is because integrating the Einstein Equations often requires complicated,
time-dependent gauge conditions
\citep{lind2007-gen_harmonic, fouc2013-compactness_and_spin}.

%\subsubsection*{Why do scattering and pair processes matter?}
Elastic scattering (see Tab.~\ref{tab:neutrino_processes}) can signicantly
modify neutrino distributions in angle and dilute the emitted spectrum over a
larger emitting surface \citep{pere2016-asl}.
This is especially pronounced in the case of heavy-lepton neutrinos.
\todo{cite}
Inelastic scattering and pair processes can introduce further modifications.
Any phenomena that involve neutrino-neutrino interactions,
for example neutrino oscillations and neutrino-antineutrino annihilation,
\todo{cite}
depend sensitively on the angular distribution.
And spectral changes can strongly affect the nucleosynthesis
occuring in the ejected and irradiated material.
\todo{justify these statements}
The ray tracing method we present in this paper captures the dominant effects
of elastic scattering, while leaving the physics of inelastic scattering and
pair processes for later work.

%\subsubsection*{In what ways is this method limited?}
Ray tracing is ideally suited to problems requiring detailed knowledge of
radiation distribution functions over small regions of spacetime:
for example along a matter or radiation trajectory,
or over a small volume outside a source.
But the method becomes computationally burdensome when the inquiry extends
to large volumes of spacetime.

More fundamentally, though, ray tracing is limited by its naive
treatment of the Boltzmann Equation (Eqn.~\ref{eqn:boltzmann}),
a treatment which essentially decouples different neutrino momenta and species.
\todo{clarify how momenta are decoupled}
When we're solving for $f^{\nu_\sigma}(x^\alpha;p_\beta)$, we have no
information about $f^{\nu_\sigma}(x^\alpha;p'_\beta)$, and
$f^{\bar{\nu}_\sigma}(x^\alpha;p'_\beta)$, which are essential ingredients
for the source terms of the Boltzmann Equation describing the creation and
destruction of neutrinos due to the interaction processes described in
Tab.~\ref{tab:neutrino_processes}.

In this paper we outwit this limitation by incorporating coupled source terms
that depend on either previously-evolved or analytical
estimates of the neutrino fields.
We also retain the freedom to drop the coupling terms, in which case our method
becomes similar to existing state-of-the-art ray tracing methods.
Our method is not a standalone radiation transport scheme,
but serves as the final component of a hybrid scheme,
piggy-backing on a lower-order radiation transport method as a
post-processing step.

\subsubsection*{Outline}
In Sec.~\ref{sec:formulation} we derive the ray tracing equations from the
Boltzmann Equation and describe our numerical scheme.
In Sec.~\ref{sec:tests} we present tests of the code.
In Sec.~\ref{sec:applications} we present several applications
in the dynamical environment following the merger of two neutron stars:
computing neutrino spectra as a function of observer's angle,
computing the self-interaction potential descriptive of flavor evolution
in high-neutrino-density environments,
and computing the energy dumped into jet-forming electron-positron
pairs via neutrino-antineutrino annihilation.

Greek tensor indices ($\alpha, \beta, ...$) range over all four coordinates,
whereas Latin indices ($i, j, ...$) range over the spatial coordinates 1--3.
We use naturalized units in which $\hbar=c=1$.
And for the remainder of this article we suppress the neutrino species label
$\nu_\sigma$ since the formulation is general to any species.

\section{Ray tracing formulation}
\label{sec:formulation}
The neutrino distribution function, $f(x^\alpha; p_\beta)$ is an invariant
quantity counting the number of neutrinos in a given six-volume of phase
space centered on $(x^\alpha,p_\beta)$.
The phase space volume elements are defined with respect to a fiducial
observer passing through event $x^\alpha$ with velocity $u^\alpha$:
\begin{align}
  \label{eqn:dV}
  dV & \equiv \sqrt{-\psi} \, dx \, dy \, dz \, u^t \\
  \label{eqn:dP}
  dP & \equiv \frac{1}{\sqrt{-\psi}} \, dp_x \, dp_y \, dp_z \,
  \frac{\varepsilon}{p^t},
\end{align}
where $\psi$ represents the determinant of the spacetime metric,
the index $t$ indicates the time-component of the given four-vector, and
\begin{equation}
  \label{eqn:varepsilon}
  \varepsilon \equiv -p_\mu u^\mu
\end{equation}
is the neutrino energy measured by our observer.
The number of particles in a given six-volume is
\begin{equation}
  dN=\frac{g}{(2\pi)^3}\,f\,dV\,dP,
\end{equation}
where $g$ counts the number of spin states accessible to the
particles ($g=1$ for neutrinos), and $f$ is the distribution function.
Each of $dV$, $dP$, $dN$, and $f$ are spacetime invariants
\citep{debb2009-gr_boltzmann_1, debb2009-gr_boltzmann_2, lind1966-gr_boltzmann}.

We may decompose the neutrino momentum like
\begin{equation}
  \label{eqn:def_momentum}
  p_\beta = \varepsilon (u_\beta + \ell_\beta),
\end{equation}
with $\ell_\beta$ the direction normal subject to the constraints
$u^\alpha \ell_\alpha = 0$ and $\ell^\alpha \ell_\alpha=1$.
With this decompositon we can write the arguments to the distribution function
$f(x^\alpha;\varepsilon,\ell_\beta)$.

Because $\ell_\beta$ is subject to two constraints
(normalization and orthogonality to the observer's velocity)
it has two degrees of freedom; we make this explicit by defining its
spatial components with respect to spherical polar angles
\begin{equation}
  \label{eqn:def_direction}
  \ell_\alpha :=
  q (s,\sin\upsilon_1\cos\upsilon_2,\sin\upsilon_1\sin\upsilon_2,\cos\upsilon_1),
\end{equation}
with $q$ and $s$ functions of $\upsilon_1$ and $\upsilon_2$.
Now our symbol for the distribution function,
$f(x^\alpha;\varepsilon,\upsilon_1,\upsilon_2)$,
makes manifest its seven independent arguments.
\todo{also decompose wrt $p_t$}

\subsection{Boltzmann Equation}
\label{ssec:boltzmann}
Neutrino radiation obeys the relativistic Boltzmann Equation,
which, suppressing the arguments of $f$ for simplicity, is written,
\todo{point out limit of QKEs}
\begin{equation}
  \label{eqn:boltzmann}
  \frac{d}{d\lambda}f = C[f],
\end{equation}
where $d/d\lambda$ denotes a derivative with respect to the affine
parameter defining the neutrino momentum (Eqn.~\ref{eqn:geodesic_x})
and $C[f]$ is the source term arising from interactions with the medium,
having dimension ${\rm energy}\,{\rm length}^{-1}$.
\todo{describe $C[f]$ in physical terms}
The source term varies over phase space $(x^\alpha,p_\beta)$,
and due to Fermi blocking depends locally on the distribution function $f$
and nonlocally on the distribution function of this neutrino and its
antiparticle at different momenta, $f'$ and $\bar{f}'$.
For simplicity we symbolize all of these dependencies with the shorthand $C[f]$.
The various neutrino interactions contributing to $C[f]$ are detailed in
App.~\ref{sec:source_terms}.

We make the right hand side of Eqn.~\ref{eqn:boltzmann} explicit by writing
the source term linear in $f$:
\begin{equation}
  \label{eqn:boltzmann_linear}
  \frac{d}{d\lambda}f =
  \mathscr{E} - \mathscr{K} f,
\end{equation}
where we have introduced
\todo{describe $\mathscr{E}$ and $\mathscr{K}$ in physical terms}
$\mathscr{E}$, the invariant total emissivity, and
$\mathscr{K}$, the invariant total opacity.
These coefficients are computed by considering their dependence
on neutrino and antineutrino distribution functions at other momenta
(i.e.\ Fermi-blocking).

We consider two classes of interactions in this work:
the absorption/emission (AE) and elastic scattering (SE) processes
listed in Tab.~\ref{tab:neutrino_processes}.
Thus we separate these coefficients:
\begin{align}
  \label{eqn:inv_emissivity_components}
  \mathscr{E} &= \mathscr{E}_{\rm AE} + \mathscr{E}_{\rm SE},\\
  \label{eqn:inv_opacity_components}
  \mathscr{K} &= \mathscr{K}_{\rm AE} + \mathscr{K}_{\rm SE}.
\end{align}

The absorption/emission coefficients are computed from sums
over the relevant emissivities and opacities
in Tab.~\ref{tab:neutrino_processes}:
\todo{define $j$ and $\chi$}
\begin{align}
  \label{eqn:ae_emissivity_summed}
  \mathscr{E}_{\rm AE}(\varepsilon)
  &= \varepsilon \sum_{i\,{\rm reactions}} j_i(\varepsilon), \\
  \label{eqn:ae_opacity_summed}
  \mathscr{K}_{\rm AE}(\varepsilon)
  &= \frac{1}{1-f^{\rm eq}(\varepsilon)} \,
  \varepsilon \sum_{i\,{\rm reactions}} \chi_{{\rm a},i}(\varepsilon),
\end{align}
where $f^{\rm eq}$ is the Fermi-Dirac distribution function with the chemical
potential dependent on the local density, temperature, and composition of the
fluid.
The appearance of $f^{\rm eq}$ in Eqn.~\ref{eqn:ae_opacity_summed} is due to
the Fermionic nature of the neutrinos, causing $\mathscr{K}_{\rm AE}$ to be
different than the simple absorption opacity,
a phenomenon called stimulated absorption.
See App.~\ref{ssec:sources_ae} for details.
\todo{identify $\mathscr{K}_{\rm AE}$ and $\kappa^*$,
  in e.g.\ \cite{hari2010-gr_nunubar_collapsar}}

The elastic scattering coefficients are computed from
a background field, and a sum over opacities for
the relevant reactions in Tab.~\ref{tab:neutrino_processes}:
\begin{align}
  \label{eqn:se_emissivity_summed}
  \mathscr{E}_{\rm SE}(\varepsilon)
  &= \mathscr{K}_{\rm SE}(\varepsilon) \, \Phi_{\rm SE}(\varepsilon) \, \\
  \label{eqn:se_opacity_summed}
  \mathscr{K}_{\rm SE}(\varepsilon)
  &= \varepsilon \sum_{i\,{\rm reactions}} \chi_{{\rm s},i}(\varepsilon).
\end{align}
See App.~\ref{ssec:sources_se} for details.

This paper introduces the use of neutrino densities and fluxes evolved in an
M1 transport simulation to estimate the background field, $\Phi_{\rm SE}$.
App.~\ref{ssec:sources_se} details a method to calculate
$\Phi_{\rm SE}(\varepsilon)$ in two different ways:
\begin{itemize}
\item
  the \emph{spectral} method using densities and fluxes
  extracted from a simulation evolved over multiple energy groups
  to compute the background field with Eqn.~\ref{eqn:se_factor_expansion},
\item
  the \emph{gray} method using energy-integrated densities and fluxes
  extracted from a gray simulation and approximating the
  energy distribution with Eqns.~\ref{eqn:J_from_gray} and
  \ref{eqn:H_from_gray}.
\end{itemize}

\subsection{Trajectories}
\label{ssec:trajectories}
Each trajectory is uniquely labeled by a pair of vectors
giving some event on the trajectory, $x^\alpha$,
and the momentum at that event, $p_\beta$.
A family of trajectories shares the same $x^\alpha$; for example,
the family of emission trajectories is ($x^\alpha_e,p_\beta$),
and the family of observation trajectories is ($x^\alpha_o,p_\beta$).

Neutrino trajectories obey the geodesic equation, which may be decomposed into
the coupled first-order equations
\begin{equation}
\label{eqn:geodesic_x}
  \frac{d x^\alpha}{d\lambda} = p^\alpha,
\end{equation}
and
\begin{equation}
\label{eqn:geodesic_p}
  \frac{d p_\beta}{d\lambda} = -\Gamma^\alpha_{\beta\gamma} p^\gamma p_\alpha,
\end{equation}
where $p^\alpha=\psi^{\alpha\beta}p_\beta$,
$\psi^{\alpha\beta}$ is the inverse of the spacetime metric $\psi_{\alpha\beta}$,
and $\Gamma^\alpha_{\beta\gamma}$ are the standard connection coefficients,
\begin{equation}
  \label{eqn:christoffel}
  \Gamma^\alpha_{\beta\gamma} =
  \frac{1}{2} \psi^{\alpha\mu}
  (\psi_{\mu\beta,\gamma} + \psi_{\mu\gamma,\beta} - \psi_{\beta\gamma,\mu}),
\end{equation}
with the comma denoting a partial derivative
$\psi_{\mu\beta,\gamma}=\partial_\gamma\,\psi_{\mu\beta}$.

Each trajectory is parameterized by affine parameter, $\lambda$, increasing in
the direction of $\ell_\beta$. We label $\lambda=\lambda_e$ at $x^\alpha_e$,
as in Fig.~\ref{fig:affine_param}.
If we multiply Eqn.~\ref{eqn:geodesic_x} by $u_\alpha \equiv dx_\alpha/ds$,
we find $ds=\varepsilon \,d\lambda$ is the proper distance traversed by the
neutrino as measured by our fiducial obsever $u^\alpha$.

\begin{figure}
  \includegraphics[width=\columnwidth]{fig-affine_parameter_label}
  \caption{Affine parameterization of a neutrino trajectory of momentum
    $p_\alpha$.
    The fiducial observer with velocity $u^\alpha$ sits at $x^\alpha_o$,
    the neutrino emission event is at $x^\alpha_e$. The affine parameter
    increases from the emission event: $\lambda_e<\lambda'<\lambda''$.}
  \label{fig:affine_param}
\end{figure}

\subsection{The Rendering Equation}
\label{ssec:rendering_eqn}
We can integrate Eqn.~\ref{eqn:boltzmann_linear} directly,
\begin{align}
  \label{eqn:rendering}
  f(\lambda,\lambda_e) &=
  f_{\rm bound}(\lambda,\lambda_e) + f_{\rm integ}(\lambda,\lambda_e), \\
  \label{eqn:rendering_fbound}
  f_{\rm bound}(\lambda,\lambda_e) &=
  f(\lambda_e) e^{-\tau(\lambda,\lambda_e)}, \\
  \label{eqn:rendering_finteg}
  f_{\rm integ}(\lambda,\lambda_e) &=
  \int_{\lambda_e}^\lambda d\lambda'\,e^{-\tau(\lambda,\lambda')}
  \mathscr{E}(\lambda'),
\end{align}
where the optical depth is defined,
\todo{rewrite this section simpler; don't give integral solution
  since we don't use}
\begin{equation}
  \label{eqn:optical_depth}
  \tau(\lambda,\lambda') \equiv -\int_{\lambda'}^\lambda
  d\lambda'' \, \mathscr{K}(\lambda''),
\end{equation}
and the parameterization conventions are depicted in
Fig.~\ref{fig:affine_param}.

We further separate the integrated distribution function into
its absorption/emission and elastic scattering components,
$f_{\rm integ} = f_{\rm AE} + f_{\rm SE}$:
\begin{align}
  \label{eqn:rendering_fae}
  f_{\rm AE}(\lambda,\lambda_e)
  &= \int_{\lambda_e}^\lambda d\lambda'\,e^{-\tau(\lambda,\lambda')}
  \mathscr{E_{\rm AE}}(\lambda'), \\
  \label{eqn:rendering_fse}
  f_{\rm SE}(\lambda,\lambda_e)
  &= \int_{\lambda_e}^\lambda d\lambda'\,e^{-\tau(\lambda,\lambda')}
  \mathscr{E_{\rm SE}}(\lambda').
\end{align}
Note that the optical depth used in both of these integrands is the total
optical depth.

\subsection{Moments of the Distribution Function}
\label{ssec:moments}
\todo{code these up}
We may take angular moments of the distribution function:
\begin{align}
  \label{eqn:J}
  J(\varepsilon) &=
  \frac{\varepsilon^3}{(2\pi)^3} \oint d\Omega' f(\varepsilon, \ell'_\beta) \\
  \label{eqn:Ha}
  H^\mu(\varepsilon) &=
  \frac{\varepsilon^3}{(2\pi)^3} \oint d\Omega' f(\varepsilon, \ell'_\beta) \ell'^\mu \\
  \label{eqn:Sab}
  S^{\mu\gamma}(\varepsilon) &=
  \frac{\varepsilon^3}{(2\pi)^3} \oint d\Omega' f(\varepsilon, \ell'_\beta) \ell'^\mu \ell'^\gamma,
\end{align}
representing specific energy density, specific energy flux, and
specific momentum flux, respectively.
Here ``specific'' refers to the quantity being integrable over neutrino energy.
\todo{relate moments to other frames?}
Integrals~\ref{eqn:J}--\ref{eqn:Sab} are performed over a solid angle in
momentum space while holding $\varepsilon$ constant:
$d\Omega \equiv d(\cos\upsilon_1)\,d\upsilon_2$.
\todo{consistent with Eqn.~\ref{eqn:def_direction}?}
We will also make use of the number density and flux defined as
$G(\varepsilon)=J(\varepsilon)/\varepsilon$
and $K^\mu(\varepsilon)=H^\mu(\varepsilon)/\varepsilon$
respectively.

The energy-integrated moments are
\begin{equation}
  \label{eqn:J_H_S_eps_integrated}
  \big\{ J,H^\mu,S^{\mu\nu} \big\} = \int_0^\infty d\varepsilon \,
  \big\{ J(\varepsilon),H^\mu(\varepsilon),S^{\mu\nu}(\varepsilon) \big\},
\end{equation}
having dimension ${\rm energy}\,{\rm length}^{-3}$.

In practice we use two possible fiducial observers to define the momentum
decomposition via Eqn.~\ref{eqn:def_momentum}: the Eulerian,
or normal observer $n_\mu=-\alpha \,\partial_\mu t$,
and the fluid, or comoving observer
$u^\mu=Wn^\mu+v_{\rm E}^\mu$.
Here $t$ is coordinate time and $\alpha$ is lapse
in the standard 3+1 decomposition of the metric
\begin{equation}
  \label{eqn:adm_metric}
  \psi_{\mu\gamma} :=
  \left(
  \begin{matrix}
    -\alpha^2 + \beta^i \beta_i  & \beta_i \\
    \beta_j                      & g_{ij}
  \end{matrix}
  \right).
\end{equation}
We have also introduced
the fluid Lorentz factor $W=-n_\mu u^\mu=\alpha u^t$,
its Eulerian velocity $v_{\rm E}^\mu=g^\mu_\lambda u^\lambda$
(distinct from its coordinate velocity $v^\mu=u^\mu/u^t$),
and the projection tensor orthogonal to the normal observer
$g^\mu_\lambda=\psi^\mu_\lambda+u^\mu u_\lambda$,
whose spatial part is the 3-metric, $g_{ij}$.

We distinguish moments computed assuming an Eulerian observer with a tilde,
e.g. $\tilde{J}$, $\tilde{H}^\mu$, $\tilde{S}^{\mu\nu}$.
Note that these three Eulerian moments are identical to the lab-frame moments
$E$, $F^\mu$, and $P^{\mu\nu}$
defined in
\cite{shib2011-truncated_moment, ocon2015-gr1d_with_nu, fouc2015-m1_nsbh}.

\subsection{Numerical Implementation}
\label{ssec:numerical}
Much of our numerical implementation is borrowed from the geodesic evolution
system described in~\cite{bohn2016-code}.
We integrate Eqns.~\ref{eqn:geodesic_x} and~\ref{eqn:geodesic_p} in the form
given by \citep{hugh1994-eh_finding},
and Eqns.~\ref{eqn:rendering} and~\ref{eqn:optical_depth}
in the form given below.
By using the time-component of Eqn.~\ref{eqn:geodesic_x} ($dt=d\lambda \, p^t$)
we may transform our integrations to coordinate time.
The coupled system of ordinary differential equations is
\begin{align}
  \label{eqn:num_x}
  \frac{dx^i}{dt} &=
  g^{ij} \frac{p_j}{p^t} - \beta^{i}\\
  \label{eqn:num_p}
  \frac{dp_i}{dt} &=
  -\alpha \alpha_{,i} p^t
  + \beta^k_{,i} p_k
  - \frac{1}{2} g^{jk}_{,i}\frac{p_j p_k}{p^t}\\
  \label{eqn:num_fae}
  \frac{df_{\rm AE}}{dt} &=
  e^{-\tau} \frac{\mathscr{E_{\rm AE}}}{p^t}\\
  \label{eqn:num_fse}
  \frac{df_{\rm SE}}{dt} &=
  e^{-\tau} \frac{\mathscr{E_{\rm SE}}}{p^t}\\
  \label{eqn:num_tau}
  \frac{d\tau}{dt} &=
  - \frac{\mathscr{K}}{p^t}.
\end{align}

We integrate each ray until we reach a terminal optical depth of
$\tau_{\rm term}$ at the earliest emission event $x_e^\alpha$.
The concept of earliest emission event is a facetious construct we use
to allow us to truncate our integration at a location along the ray where any
further additions to the field are negligible due to the large optical depth
between $x_e^\alpha$ and $x_o^\alpha$.
We choose $\tau_{\rm term}$ large so that
$f_{\rm bound}$ is negligible
(in practice $e^{-\tau_{\rm term}}=0.001$)
and compute $f_{\rm bound}$ assuming radiative equilibrium with the matter,
i.e. $f_{\rm bound}=(1+\exp((\varepsilon-\mu_\nu)/T)$,
and $\mu_{\nu_e}=-\mu_{\bar{\nu}_e}=\mu_{e^{-}}+\mu_p-\mu_n$ and
$\mu_{\nu_x}=0$.
If the ray terminates off the numerical domain before the terminal optical
depth accumulates, we set $f_{\rm bound}=0$.

Since we don't know the emission event a priori,
\todo{make clear ``emission event'' doesn't involve physical assumption}
we follow the integration backwards in time,
from $t_o$ to $t_e$.
We begin each integration by setting initial values for the variables at $t_o$.
The observer specifies $x^i_o$ and $p_{i,o}$,
and we set $f_{{\rm AE},o}=f_{{\rm SE},o}=0$ and $\tau_o=0$.

We integrate Eqns.~\ref{eqn:num_x}--\ref{eqn:num_tau} with adaptive step sizes,
using the 5th-order Dormand-Prince algorithm \cite{pres2007-nr_3rd_ed},
which produces an error estimate by comparing the 4th and 5th order solutions.
\todo{confirm}
After each step is taken, the errors for each of the 9 variables of
Eqns.~\ref{eqn:num_x}--\ref{eqn:num_tau} are compared.
If the maximum of these errors exceeds some absolute or relative threshold
\todo{specify ScaledAbsRel algorithm}
the step size is decreased and the step recomputed,
if it is below some threshold the next step size is increased.
% see OdeIntegrators::AdaptiveDense
% and OdeSteppers::DormandPrince5
% and OdeControllers::ProportionalIntegral_Base::StepSizeToAttempt
% and OdeErrorMeasures::ScaledAbsRel
\todo{specify}

The volume data are components of the metric $\psi_{\alpha\beta}$,
its derivatives $\psi_{\alpha\beta,\gamma}$,
and the hydrodynamical data $W$, $u_i$, $\varrho$, $T$, and $Y_e$.
These fields are stored in spectral representation.
If they are computed via hydrodynamical simulation, they are saved to disk
at specified coordinate times and interpolated
with spectral interpolation in space,
and 1st-order polynomial interpolation in time
(as described in \cite[App.~B]{bohn2016-code}).
If computed from a quasi-stationary solution to the
\todo{clarify quasistationary timescale}
general relativistic hydrodynamics equations,
a single time is stored to disk, and no time interpolation is needed.

\section{Code Tests}
\label{sec:tests}

To test our algorithm, we integrate the ray tracing equations
(Eqns.~\ref{eqn:num_x}--\ref{eqn:num_tau}) at one observation position
in the following configurations. This suite of configurations
may be understood as a hierarchy of increasing physical realism:
beginning with a homogeneous medium of effectively
infinite extent (i.e.\ optically thick) and progressing to a model of a
1D pre-supernova, post-bounce collapse profile
evolved using an M1 transport hydrodynamical simulation.
In the pre-supernova model, we compare ray tracing distributions
to those calculated in a Monte Carlo transport simulation.
Finally, we calculate global measures of the neutrino fields outside of the
hypermassive neutron star and disk formed in a binary neutron star
merger simulated by \cite{fouc2016-m1_nsns, fouc2016-m1_evolve_n},
and compare ray tracing results to those from the M1 transport simulation.

\subsection{Infinite Homogeneous Slab:
  Testing Thermodynamic Equilibrium}
\label{ssec:test_equilibrium}
In an optically thick region the radiation field is in thermodynamic equilibrium
with the matter.
We set up a large slab of matter in two homogeneous configurations
representative of the fluid in the Massive Collapsed Star test
(Sec.~\ref{ssec:test_collapse}) at two different radii:
at 50~km
For comparision with Collapsed Star test, we use the LS180 equation of state,
in which the equilibrium
neutrino chemical potentials are
$\mu_{\nu_e}=-0.639$~MeV, and 
$\mu_{\nu_e}=5.771$~MeV respectively
($\mu_{\bar{\nu}_e}=-\mu_{\nu_e}$ and $\mu_{\nu_x}=0$).
Plot in Fig.~\ref{fig:homogeneous_isotropic}.
\todo{run test at this thermodynamic point}

>>>>

\documentclass[aps,prd,twocolumn,superscriptaddress,groupedaddress]{revtex4}
% Intended to be built with pdflatex, using REVTEX 4
% Template taken from FermiLab APS Jouranl template.
% Labeling convention example: \label{ssec:acknowledge}.
%   Specifically, use (<=4)-letter class (e.g. chap, sec, ssec, fig, etc.),
%   followed by a colon, followed by a shortish underscore separated name
%   (e.g. intro, nu_energy_spectrum).

% packages
\usepackage{amssymb}              % for \gtrsim ...
\usepackage{amsmath}              % for \text, align, split ...
\usepackage{graphicx}             % for \includegraphics
\usepackage{hyperref}             % for \url and \href

% commands/initialization
\hyphenation{ALPGEN} % avoid incorrect hyphenation
\hyphenation{EVTGEN} % ''
\hyphenation{PYTHIA} % ''

@@

% The following information is for internal review, please remove them for submission
\widetext
\leftline{Compiled \today}
\leftline{To be submitted to PRD}

\title{Neutrino Annihilation Around Massive Nuclear Accretion Disks}

\author{M.\ Brett Deaton}
\affiliation{Department of Physics \& Astronomy,
  Washington State University, Pullman, WA 99164, USA}
\affiliation{Joint Institute for Nuclear Astrophysics,
  Michigan State University, East Lansing, MI 48824, USA}
\affiliation{Department of Physics,
  North Carolina State University, Raleigh, NC 27695, USA}
\email{mbdeaton@ncsu.edu}

@@

\begin{abstract}
  This is our abstract.
\end{abstract}

@@

This is some text in the body.
The high disk mass model is described in \cite{deat2013-leakage}.

\bibliography{references}
